Do syntactic and semantic similarity lead to interference effects?\\ Evidence from self-paced reading and event-related potentials using German

### 2024-12-02 paper/revision2/elsarticle-template-harv.tex
Updated via the Open Science Framework

Semantic interference is manipulated via the animacy of the distractor. During the retrieval of the subject of the verb \textit{trank} (`drank'), the animate distractor  \textit{Einbrecher} (`burglar') is assumed to cause high semantic interference, while the inanimate distractor \textit{Verlust} (`loss') is assumed to cause low semantic interference. {The animacy of the distractor is used as a stand-in feature for semantic features relevant for the specific retrieval event which might be more fine-grained. In \ref{ex:materials}, the semantic retrieval cue might be \{+animate\} or \{+human\} but it might as well be \{+can drink\}. While more generic features like \{+animate\} or \{+human\} (all animate distractors were human) might be created during memory encoding, more specific features like \{+can drink\} are probably only activated when relevant. The present study is not designed to provide insight into the specific semantic features used during memory encoding and retrieval. For experimental work on highly specific retrieval cues like \{+sailable\} and \{+shatterable\}, see \cite{vandyke_mcelree06} and \cite{cunnings_sturt_2018retrieval}. For a principled approach to feature selection using word embeddings, see \cite{smith2020principled}. A plausibility norming experiment (see page \pageref{plausib_normin_orig}) was conducted to ensure that the semantic manipulation did work, i.e., that the semantic requirements of the verb matched the features of the subject and the animate distractor, but not the features of the inanimate distractor.
}\label{semantic_features_encoding}

@@

Alternately, retrieval interference -- especially syntactic retrieval interference -- might modulate the P600. The P600 is a positive deflection in the ERP wave form that is typically observed at least 600 ms after stimulus onset. It is associated with syntactic processing, e.g., morpho-syntactic errors and reanalysis \citep[see, e.g.,][]{osterhout&holcomb_1992, kaan_etal_2000_P600}. High interference could increase difficulty during syntactic integration of subject and verb in our materials. Potential misretrievals of the distractor might lead to reanalysis which could manifest as an increased P600 amplitude in high compared to low interference conditions.} \label{why_n400}

@@

The SPR study, hereafter Experiment 1a, was initially designed so that participants read \textcolor{blue}{half of the items (1-60) in the first session and the other half of the items (61-120) in the second session on another day.} The second session showed an adaptation effect \citep{prasad2021rapid}: Reading times decreased overall. The adaptation effect is theoretically interesting per se \citep{fine2013rapid} and we discuss it below; but it was problematic for our design because the average effect would then be confounded by adaptation. For this reason, we ran a second version of the SPR study, hereafter Experiment 1b, in which participants (N=570) were either shown items 1-60 or items 61-120 (but not both). In the Results section, we present the pooled data (N=774) from E1a and E1b.
As mentioned above, because the ERP method requires a large number of items, all the ERP data come from participants who completed both sessions.

@@

In Experiment 1a, participants were re-invited for session 2 after they completed session 1. The experimental sessions were separated by 1 - 20 days. The procedure of the sessions was identical. In session 1, the experimental items 1-60 were presented interspersed with 40 filler sentences. In session 2, the experimental items 61-120 were presented with another set of 40 filler sentences. {\Copy{fillers}{\textcolor{blue}{The fillers were included to mask the critical manipulation. They} were less syntactically complex than the experimental items but had at least one embedded clause and generally provided more variety\textcolor{blue}{, e.g., by including modifiers and different types of subordinate clauses} (e.g., ``Der Teppichmacher, der früh in die Werkstatt gekommen war, reparierte den besonders schönen, alten Teppich während er die Nachrichten hörte'', `The carpet maker who came to the workshop early repaired the especially beautiful old carpet while listening to the news.'; ``Der Jugendliche war genervt, weil seine Freundin, die nicht studieren wollte, oft die Schule schwänzte, um Computer zu spielen.'', `The teenager was annoyed because his girlfriend who did not want to go to University  skipped school frequently to play video games.').}}\label{fillers} Each session lasted approximately 25 minutes.

@@

\begin{sidewaysfigure}[h]
    \caption{Self-paced reading times with 95\% confidence intervals. Panel A and B show the pooled reading times across the whole sentence; separately for high (A) and low (B) syntactic interference due to differing sentence structure. \textcolor{blue}{Panel C shows the reading times of the sentence focusing on the regions between the distractor and critical verb for all conditions.}}
    \label{fig:whole_sentence}
    \centering
    \includegraphics[width=\textwidth]{Pandora_all_wholesentence_pooled_zoom.jpg}
\end{sidewaysfigure}
\clearpage

\label{revised_SPR}\Copy{revised_SPR}{
It is apparent from Figure \ref{fig:whole_sentence} that, regardless of the syntactic manipulation, at the distractor, reading times between high and low semantic interference started to differ: distractors in the high semantic interference conditions induced longer reading times than in the low semantic interference conditions. \textcolor{blue}{To anticipate our discussion of the SPR results here, we believe that the difference in reading times starting at the distractor was caused by encoding interference \citep{Oberauer_Kliegl_2006}. Additionally, the second session in Experiment 1a} showed shorter reading times compared to the first session. As mentioned earlier, this attenuation in reading times was likely due to adaptation to the SPR task in the second session.

@@

\textcolor{blue}{Since the reading times started to differ long before the critical verb which is assumed to be the locus of retrieval in the cue-based retrieval framework,  retrieval interference cannot be the cause of this difference. T}he most likely explanation is encoding interference \cite[e.g.,][but see the potential confounds discussed above]{Oberauer_Kliegl_2006,smith2021encoding}. The increased effort of encoding and subsequently maintaining representations of three different animate noun phrases (the introduction noun, the subject and the animate distractor) is the most likely reason for the slowdown in high vs.\ low semantic interference conditions \citep[for similar findings, see e.g., ][]{lago_etal_2021, ness2019, ness2017, kush_etal_2015, gordon02}.

@@

\textcolor{blue}{In summary, the main finding of the Experiment 1 is that the reading times were primarily affected by semantic encoding interference.} We turn next to the event-related potentials experiment.

>>>>

Semantic interference is manipulated via the animacy of the distractor. During the retrieval of the subject of the verb \textit{trank} (`drank'), the animate distractor  \textit{Einbrecher} (`burglar') is assumed to cause high semantic interference, while the inanimate distractor \textit{Verlust} (`loss') is assumed to cause low semantic interference. {The animacy of the distractor is used as a stand-in feature for semantic features relevant for the specific retrieval event which might be more fine-grained. In \ref{ex:materials}, the semantic retrieval cue might be \{+animate\} or \{+human\} but it might as well be \{+can drink\}. While more generic features like \{+animate\} or \{+human\} (all animate distractors were human) might be created during memory encoding, more specific features like \{+can drink\} are probably only activated when relevant. The present study is not designed to provide insight into the specific semantic features used during memory encoding and retrieval. For experimental work on highly specific retrieval cues like \{+sailable\} and \{+shatterable\}, see \cite{vandyke_mcelree06} and \cite{cunnings_sturt_2018retrieval}. For a principled approach to feature selection using word embeddings, see \cite{smith2020principled}. A plausibility norming experiment (see below) was conducted to ensure that the semantic manipulation did work, i.e., that the semantic requirements of the verb matched the features of the subject and the animate distractor, but not the features of the inanimate distractor.
}\label{semantic_features_encoding}

@@

Alternatively, retrieval interference -- especially syntactic retrieval interference -- might modulate the P600. The P600 is a positive deflection in the ERP wave form that is typically observed at least 600 ms after stimulus onset. It is associated with syntactic processing, e.g., morpho-syntactic errors and reanalysis \citep[see, e.g.,][]{osterhout&holcomb_1992, kaan_etal_2000_P600}. High interference could increase difficulty during syntactic integration of subject and verb in our materials. Potential misretrievals of the distractor might lead to reanalysis which could manifest as an increased P600 amplitude in high compared to low interference conditions.} \label{why_n400}

@@

The SPR study, hereafter Experiment 1a, was initially designed so that participants read \textcolor{blue}{half of the items (1-60) in the first session and the other half of the items (61-120) in the second session on another day.} The second session showed an adaptation effect \citep{prasad2021rapid}: Reading times decreased overall. The adaptation effect is theoretically interesting per se \citep{fine2013rapid}; but it was problematic for our design because the average effect would then be confounded by adaptation. For this reason, we ran a second version of the SPR study, hereafter Experiment 1b, in which participants (N=570) were either shown items 1-60 or items 61-120 (but not both). In the Results section, we present the pooled data (N=774) from E1a and E1b.
As mentioned above, because the ERP method requires a large number of items, all the ERP data come from participants who completed both sessions.

@@

In Experiment 1a, participants were re-invited for session 2 after they completed session 1. The experimental sessions were separated by 1 - 20 days. The procedure of the sessions was identical. In session 1, the experimental items 1-60 were presented interspersed with 40 filler sentences. In session 2, the experimental items 61-120 were presented with another set of 40 filler sentences. {\Copy{fillers}{\textcolor{blue}{The fillers were included to mask the critical manipulation. They} were less syntactically complex than the experimental items but had at least one embedded clause and generally provided more variety\textcolor{blue}{, e.g., by including modifiers and different types of subordinate clauses} (e.g., ``Der Teppichmacher, der früh in die Werkstatt gekommen war, reparierte den besonders schönen, alten Teppich, während er die Nachrichten hörte'', `The carpet maker who came to the workshop early repaired the especially beautiful old carpet while listening to the news.'; ``Der Jugendliche war genervt, weil seine Freundin, die nicht studieren wollte, oft die Schule schwänzte, um Computer zu spielen.'', `The teenager was annoyed because his girlfriend who did not want to go to University  skipped school frequently to play video games.').}}\label{fillers} Each session lasted approximately 25 minutes.

@@

\begin{sidewaysfigure}[h]
    \caption{Self-paced reading times with 95\% confidence intervals. Panel A and B show the pooled reading times across the whole sentence; separately for high (A) and low (B) syntactic interference due to differing sentence structure. \textcolor{blue}{Panel C shows the reading times of the sentence focusing on the regions between the distractor and critical verb for all conditions.}}
    \label{fig:whole_sentence}
    \centering
    \includegraphics[width=\textwidth]{Pandora_all_wholesentence_pooled.jpg}
\end{sidewaysfigure}
\clearpage

\label{revised_SPR}\Copy{revised_SPR}{
It is apparent from Figure \ref{fig:whole_sentence} that, regardless of the syntactic manipulation, at the distractor, reading times between high and low semantic interference started to differ: distractors in the high semantic interference conditions induced longer reading times than in the low semantic interference conditions. \Copy{smith1}{\textcolor{blue}{To anticipate our discussion of the SPR results here, we believe that the difference in reading times starting at the distractor was caused by encoding interference \citep{Oberauer_Kliegl_2006, smith2021encoding}}}\label{smith1}. \textcolor{blue}{Additionally, the second session in Experiment 1a} showed shorter reading times compared to the first session. As mentioned earlier, this attenuation in reading times was likely due to adaptation to the SPR task in the second session.

@@

\textcolor{blue}{Since the reading times started to differ long before the critical verb which is assumed to be the locus of retrieval in the cue-based retrieval framework,  retrieval interference cannot be the cause of this difference.} \Copy{smith2}{The most likely explanation is encoding interference \cite[e.g.,][but see the potential confounds discussed above]{Oberauer_Kliegl_2006,smith2021encoding}.}\label{smith2} The increased effort of encoding and subsequently maintaining representations of three different animate noun phrases (the introduction noun, the subject and the animate distractor) is the most likely reason for the slowdown in high vs.\ low semantic interference conditions \citep[for similar findings, see e.g., ][]{lago_etal_2021, ness2019, ness2017, kush_etal_2015, gordon02}.

@@

\textcolor{blue}{In summary, the main finding of Experiment 1 is that the reading times were primarily affected by semantic encoding interference.} We turn next to the event-related potentials experiment.
